---
title: "Heatmaps"
output: html_document
date: "2024-12-07"
---


```{r}
library(pheatmap)
library(dplyr)
library(caret)
library(MLeval)
library(pROC)
library(h2o)
library(edgeR)
library(pROC)
library(pheatmap)
library(RColorBrewer)
library("FactoMineR")
library("devtools")
library("factoextra")
library(cluster)

h2o.init(nthreads = -1) 

```



```{r}
mel_dat <- readRDS("validation_lcpm.rds")

l_dat <- readRDS("skcm_ready.RDS")

l_dat$meta.multiple.bin <- as.factor(l_dat$meta.multiple.bin)

l_dat <- l_dat %>% select(-meta.multiple.gene)

l_dat <- l_dat %>% select(intersect(colnames(l_dat), colnames(mel_dat)))


library(caret)

set.seed(3456)

trainIndex <- createDataPartition(l_dat$meta.multiple.bin, p = .7, 
                                  list = FALSE, 
                                  times = 1)

train <- l_dat[ trainIndex,]

test <- l_dat[-trainIndex,]

summary(train$meta.multiple.bin)

summary(test$meta.multiple.bin)



```


```{r}
common_cols <- intersect(colnames(l_dat), colnames(mel_dat))

library(msigdbr)

infg_gene_sets = msigdbr(species = "human", category = "H")

infg_gene_sets = infg_gene_sets %>%  filter(gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE")

sum(duplicated(infg_gene_sets$gene_symbol))

infg_gene_sets = infg_gene_sets[!duplicated(infg_gene_sets$gene_symbol), ]

infg_gene_set = infg_gene_sets$gene_symbol

infg_gene_set <- gsub("\\-| ", ".", infg_gene_set)

sum(infg_gene_set %in% colnames(mel_dat))  

infg_gene_set = intersect(infg_gene_set, common_cols)

infg_gene_set = append(infg_gene_set, "meta.multiple.bin")
```

```{r}

ifng_gene_exp_tr <- select(train, intersect(colnames(train), infg_gene_set))
ifng_gene_exp_tr <- ifng_gene_exp_tr[order(ifng_gene_exp_tr$meta.multiple.bin), ]
ifng_gene_exp_tr$meta.data <- "train"

```

```{r}

ifng_gene_exp_te <- select(test, intersect(colnames(test), infg_gene_set))
ifng_gene_exp_te <- ifng_gene_exp_te[order(ifng_gene_exp_te$meta.multiple.bin), ]
ifng_gene_exp_te$meta.data <- "test"

```

# Expression Heatmap

```{r}
library(RColorBrewer)

ifng_gene_exp <- rbind(ifng_gene_exp_te, ifng_gene_exp_tr)
ann_row = select(ifng_gene_exp, starts_with("meta."))

ann_row = rename(ann_row, data = meta.data)
ann_row = rename(ann_row, categorized = meta.multiple.bin)

ifng_gene_count = select(ifng_gene_exp, -starts_with("meta."))

ann_colors = list(
  data = c("test" = "#B15928" , "train" = "#33A02C"),
  categorized = c( "High" = "#E31A1C", "Low" = "#6A3D9A")
)


# Create custom column labels
col_labels <- colnames(ifng_gene_count)

seq = seq(1, 188, by = 14)

col_labels <- ifelse(col_labels %in% col_labels[seq] == 0, "", col_labels )


pheatmap(t(ifng_gene_count), annotation_col  = ann_row, show_colnames=F,scale = "row", annotation_colors = ann_colors,labels_row =col_labels, fontsize_row = 10, cluster_cols=F, col = rev(brewer.pal(11, "RdBu")), cluster_rows =F, show_rownames = F)

```




# model loading

```{r, echo=FALSE}
xgb_model <- readRDS("./all_models/xgb_valid_final.RDS")
rf_model<- readRDS("./all_models/rf_valid_final.RDS")
glmnet_model <- readRDS("./all_models/glmnet_valid_final.RDS")
ann_model_1 <- h2o.loadModel("C:\\EKIZ_LAB\\all_models\\ann_valid_fi.RDS\\dl_grid_no_model_4")

## diff expr.
train_diff_genes <- readRDS("TCGA_SKCM_train_de_genes_IFNG_final.rds") 
test_diff_genes <- readRDS("TCGA_SKCM_test_de_genes_IFNG_final.rds")  
```


# important gene extraction

```{r}

#xgboost

xgbImp <- varImp(xgb_model, scale = T)
xgbImp_d <- data.frame(xgbImp$importance)
xgbImp_d <- xgbImp_d %>% filter(!Overall == 0)
xgbImp_d$gene <- row.names(xgbImp_d)
colnames(xgbImp_d)[colnames(xgbImp_d) == "Overall"] ="xgb_score"



# random forest

rfImp <- varImp(rf_model, scale = T)
rfImp_d <- data.frame(rfImp$importance)
rfImp_d <- rfImp_d  %>% filter(!Overall == 0)
rfImp_d$gene <- rownames(rfImp_d)
colnames(rfImp_d)[colnames(rfImp_d) == "Overall"] ="rf_score"
# rfImp_d$rf_score <- signif(rfImp_d$rf_score, digits  = 8) 

# glmnet

glmnetImp <- varImp(glmnet_model, scale = T)
glmnetImp_d <- data.frame(glmnetImp$importance)
glmnetImp_d <- glmnetImp_d %>% filter(!Overall == 0)
glmnetImp_d$gene <- rownames(glmnetImp_d)
colnames(glmnetImp_d)[colnames(glmnetImp_d) == "Overall"] ="glmnet_score"

# ann 

ann_varimp_1 <- h2o.varimp(ann_model_1)
ann_varimp_1_d <- as.data.frame(ann_varimp_1)
# identical(ann_varimp_1_d$relative_mportance, ann_varimp_1_d$scaled_mportance)
# [1] TRUE
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "variable"] ="gene"
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "scaled_mportance"] ="ann_score"
ann_varimp_1_d <- ann_varimp_1_d %>% select(gene, ann_score)

# diff expr.

test_diff_genes <- test_diff_genes %>% 
                   filter(padj < 0.05) %>% 
                   mutate(wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, wlfc)


train_diff_genes <-train_diff_genes %>% 
                   filter(padj < 0.05) %>% 
                   mutate(wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, wlfc)

colnames(test_diff_genes) <- paste("test", colnames(test_diff_genes), sep = "_")
colnames(train_diff_genes) <- paste("train", colnames(train_diff_genes), sep = "_")

colnames(test_diff_genes)[colnames(test_diff_genes) == "test_gene"] ="gene"
colnames(train_diff_genes)[colnames(train_diff_genes) == "train_gene"] ="gene"


```


```{r}
common_gene_list<- list(
  ann = ann_varimp_1_d$gene,
  xgb = xgbImp_d$gene,
  glmnet = glmnetImp_d$gene,
  train_diff_genes= test_diff_genes$gene,
  test_diff_genes = train_diff_genes $gene,
  rf = rfImp_d$gene
)

common_elements <- Reduce(intersect, common_gene_list)

```


# merging for distance

```{r}

v1 <- full_join(ann_varimp_1_d, rfImp_d, by= "gene") 
v2 <- full_join(xgbImp_d,glmnetImp_d, by= "gene")
v3 <- full_join(train_diff_genes,test_diff_genes, by= "gene")
v4 <- full_join(v1,v2, by= "gene")
v5 <- full_join(v3,v4, by= "gene")

merged_genes <- v5

rownames(merged_genes) <- merged_genes$gene

merged_genes<- merged_genes %>% na.omit(merged_genes)

merged_genes <- select(merged_genes, -gene)

```

# Heatmap for genes using non-normalized data

```{r}

# distance_matrix_scaled <- as.matrix(get_dist(merged_genes, method = "canberra")) 


pheatmap(merged_genes,
         main = "Heatmaps for Common Genes",
         scale = "row",
         # display_numbers = TRUE,
         # fontsize_number = 10, 
         treeheight_row = 0, treeheight_col = 0,

         color = rev(brewer.pal(n= 20, "RdBu"))
)

```






# Canberra Distance Matrix (No Zero No Filtering)

```{r,fig.height= 4, fig.width=5.5}

distance_matrix_scaled <- as.matrix(get_dist(t(merged_genes), method = "canberra")) 


pheatmap(distance_matrix_scaled,
         main = "Canberra Distance Matrix",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 20, "RdBu"))
)
```

# Kendall Distance Matrix (No Zero No Filtering)

```{r,fig.height= 4, fig.width=5.5}
w_mat <- cor(merged_genes, method = 'kendall')

pheatmap(w_mat,
         main = "Kendall Matrix",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 
         # annotation_col = alg_col,
         # annotation_row = alg_col,
         color = rev(brewer.pal(n= 9, "RdBu"))
)
```



```{r}
preprocessed_data <- preProcess(merged_genes, method = 'range')
transformed_data <- predict(preprocessed_data, merged_genes)
```

# Canberra Distance Matrix (No Zero No Filtering Normalized)

```{r,fig.height= 4, fig.width=5.5}
library(pheatmap)
library(RColorBrewer)
library("FactoMineR")
library("devtools")
library("factoextra")
library(cluster)

distance_matrix_scaled <- as.matrix(get_dist(t(transformed_data), method = "canberra")) 


pheatmap(distance_matrix_scaled,
         main = "             Canberra Distance Matrix ",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdBu"))
)
```





# Kendall Distance Matrix (No Zero No Filtering Normalized)

```{r,fig.height= 4, fig.width=5.5}
w_mat <- cor(transformed_data, method = 'kendall')

pheatmap(w_mat,
         main = "Kendall Matrix ",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 
         # annotation_col = alg_col,
         # annotation_row = alg_col,
         color = rev(brewer.pal(n= 9, "RdBu"))
)
```


# Euclidean (No Zero No Filtering Normalized)
```{r, fig.height= 4, fig.width=5.5}
distance_matrix_scaled <- as.matrix(get_dist(t(transformed_data), method = "euclidean")) 


pheatmap(distance_matrix_scaled,
         main = "       Euclidean Distance Matrix",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdBu"))
)
```

# Heatmap for genes using normalized data 

```{r}

# distance_matrix_scaled <- as.matrix(get_dist(transformed_data, method = "canberra")) 


pheatmap(transformed_data,
         scale = "row",
         main = " Heatmap for Common Genes ",
         # display_numbers = TRUE,
         # fontsize_number = 10, 
         treeheight_row = 0, treeheight_col = 0,
         color = rev(brewer.pal(n= 9, "RdBu"))
)

```





```{r}
## diff expr.
train_diff_genes <- readRDS("TCGA_SKCM_train_de_genes_IFNG_final.rds") 
test_diff_genes <- readRDS("TCGA_SKCM_test_de_genes_IFNG_final.rds")  
```



```{r}
#xgboost

xgbImp <- varImp(xgb_model, scale = T)
xgbImp_d <- data.frame(xgbImp$importance)
xgbImp_d <- xgbImp_d %>% filter(!Overall == 0)
xgbImp_d$gene <- row.names(xgbImp_d)
colnames(xgbImp_d)[colnames(xgbImp_d) == "Overall"] ="xgb_score"



# random forest

rfImp <- varImp(rf_model, scale = T)
rfImp_d <- data.frame(rfImp$importance)
rfImp_d <- rfImp_d  %>% filter(!Overall == 0)
rfImp_d$gene <- rownames(rfImp_d)
colnames(rfImp_d)[colnames(rfImp_d) == "Overall"] ="rf_score"
# rfImp_d$rf_score <- signif(rfImp_d$rf_score, digits  = 8) 

# glmnet

glmnetImp <- varImp(glmnet_model, scale = T)
glmnetImp_d <- data.frame(glmnetImp$importance)
glmnetImp_d <- glmnetImp_d %>% filter(!Overall == 0)
glmnetImp_d$gene <- rownames(glmnetImp_d)
colnames(glmnetImp_d)[colnames(glmnetImp_d) == "Overall"] ="glmnet_score"

# ann 

ann_varimp_1 <- h2o.varimp(ann_model_1)
ann_varimp_1_d <- as.data.frame(ann_varimp_1)
# identical(ann_varimp_1_d$relative_mportance, ann_varimp_1_d$scaled_mportance)
# [1] TRUE
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "variable"] ="gene"
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "scaled_mportance"] ="ann_score"
ann_varimp_1_d <- ann_varimp_1_d %>% select(gene, ann_score)

# diff expr.

test_diff_genes$significance<- "NO"
test_diff_genes$significance[test_diff_genes$padj < 0.05 & test_diff_genes$log2FoldChange > 1 ] <- "UP"
test_diff_genes$significance[test_diff_genes$padj < 0.05 & test_diff_genes$log2FoldChange < -1 ] <- "DOWN"

train_diff_genes$significance<- "NO"
train_diff_genes$significance[train_diff_genes$padj < 0.05 & train_diff_genes$log2FoldChange > 1 ] <- "UP"
train_diff_genes$significance[train_diff_genes$padj < 0.05 & train_diff_genes$log2FoldChange < -1 ] <- "DOWN"





test_diff_genes <- test_diff_genes %>% 
                   filter(significance != "NO") %>% 
                   mutate(test_wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, test_wlfc)


train_diff_genes <-train_diff_genes %>% 
                   filter(significance != "NO") %>% 
                   mutate(train_wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, train_wlfc)

```



# merging for distance

```{r}

v1 <- full_join(ann_varimp_1_d, rfImp_d, by= "gene") 
v2 <- full_join(xgbImp_d,glmnetImp_d, by= "gene")
v3 <- full_join(train_diff_genes,test_diff_genes, by= "gene")
v4 <- full_join(v1,v2, by= "gene")
v5 <- full_join(v3,v4, by= "gene")

merged_genes <- v5

rownames(merged_genes) <- merged_genes$gene

merged_genes<- merged_genes %>% na.omit(merged_genes)

merged_genes <- select(merged_genes, -gene)

```


# Canberra Distance Matrix (No Zero Yes Filtering)

```{r,fig.height= 4, fig.width=5.5}
library(pheatmap)
library(RColorBrewer)
library("FactoMineR")
library("devtools")
library("factoextra")
library(cluster)

distance_matrix_scaled <- as.matrix(get_dist(t(merged_genes), method = "canberra")) 


pheatmap(distance_matrix_scaled,
         main = "             Canberra Distance Matrix",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdBu"))
)
```


# Heatmap for genes using non-normalized data

```{r}

# distance_matrix_scaled <- as.matrix(get_dist(merged_genes, method = "canberra")) 


pheatmap(merged_genes,
         scale = "row",
         main = "Heatmaps for Common Genes",
         # display_numbers = TRUE,
         # fontsize_number = 10, 
         treeheight_row = 0, treeheight_col = 0,

         color = rev(brewer.pal(n= 9, "RdBu"))
)

```


# Kendall Distance Matrix (No Zero Yes Filtering)

```{r,fig.height= 4, fig.width=5.5}
w_mat <- cor(merged_genes, method = 'kendall')

pheatmap(w_mat,
         main = "Kendall Matrix ",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 
         # annotation_col = alg_col,
         # annotation_row = alg_col,
         color = rev(brewer.pal(n= 9, "RdBu"))
)
```


```{r}
preprocessed_data <- preProcess(merged_genes, method = 'range')
transformed_data <- predict(preprocessed_data, merged_genes)
```

# Canberra Distance Matrix (No Zero Yes Filtering Normalized)

```{r,fig.height= 4, fig.width=5.5}
library(pheatmap)
library(RColorBrewer)
library("FactoMineR")
library("devtools")
library("factoextra")
library(cluster)

distance_matrix_scaled <- as.matrix(get_dist(t(transformed_data), method = "canberra")) 


pheatmap(distance_matrix_scaled,
         main = "Canberra Distance Matrix ",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdBu"))
)
```

# Kendall Distance Matrix (No Zero Yes Filtering Normalized)

```{r,fig.height= 4, fig.width=5.5}
w_mat <- cor(transformed_data, method = 'kendall')

pheatmap(w_mat,
         main = "Kendall Matrix ",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 
         # annotation_col = alg_col,
         # annotation_row = alg_col,
         color = rev(brewer.pal(n= 9, "RdBu"))
)
```


# Euclidean (No Zero Yes Filtering Normalized)
```{r, fig.height= 4, fig.width=5.5}
distance_matrix_scaled <- as.matrix(get_dist(t(transformed_data), method = "euclidean"))


pheatmap(distance_matrix_scaled,
         main = " Euclidean Distance Matrix",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdBu"))
)
```


# Heatmap for genes (using normalized data)

```{r}

# distance_matrix_scaled <- as.matrix(get_dist(transformed_data, method = "canberra")) 


pheatmap(transformed_data,
         scale = "row",
         main = "Heatmaps for Common Genes ",
         treeheight_row = 0, treeheight_col = 0,
         color = rev(brewer.pal(n= 9, "RdBu"))
)

```

