---
title: "Outlier_Analysis"
author: Arda Askin
date: "`r Sys.Date()`"   
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    theme: sandstone
--- 
# package loading

```{r, echo=FALSE, , message=FALSE, , warning=FALSE}
library(dplyr)
library(caret)
library(MLeval)
library(pROC)
library(h2o)
library(edgeR)
library(pROC)

h2o.init(nthreads = -1) 

```

# model loading

```{r, echo=FALSE}
xgb_model <- readRDS("./all_models/xgb_valid_final.RDS")
rf_model<- readRDS("./all_models/rf_valid_final.RDS")
glmnet_model <- readRDS("./all_models/glmnet_valid_final.RDS")
ann_model_1 <- h2o.loadModel("C:\\EKIZ_LAB\\all_models\\ann_valid_fi.RDS\\dl_grid_no_model_4")




## diff expr.
train_diff_genes <- readRDS("TCGA_SKCM_train_de_genes_IFNG_final.rds") 
test_diff_genes <- readRDS("TCGA_SKCM_test_de_genes_IFNG_final.rds")  

```


#model hyperparameters

```{r}

rf_model$finalModel$ntree
rf_model$finalModel$mtry
xgb_model$finalModel$params
glmnet_model$finalModel$tuneValue

deep_params <- ann_model_1@parameters

params_dt <- as.data.frame(as.list(deep_params))

params_dt <- t(params_dt[1, ])

colnames(params_dt)[1] <- "Value" 

params_dt

```



# Data Prep
```{r}
mel_dat <- readRDS("validation_lcpm.rds")

l_dat <- readRDS("skcm_ready.RDS")

l_dat$meta.multiple.bin <- as.factor(l_dat$meta.multiple.bin)

l_dat <- l_dat %>% select(-meta.multiple.gene)

l_dat <- l_dat %>% select(intersect(colnames(l_dat), colnames(mel_dat)))


library(caret)

set.seed(3456)

trainIndex <- createDataPartition(l_dat$meta.multiple.bin, p = .7, 
                                  list = FALSE, 
                                  times = 1)

train <- l_dat[ trainIndex,]

test <- l_dat[-trainIndex,]

summary(train$meta.multiple.bin)

summary(test$meta.multiple.bin)


l_tr <- train %>% select(-starts_with("meta."))
l_te <- test %>% select(-starts_with("meta."))

mel_dat <- mel_dat %>% select(intersect(colnames(l_dat), colnames(mel_dat)))

l_te_2 <- mel_dat %>% select(-starts_with("meta."))

```
# Validation Data AUC-ROC Plot

```{r}
 
test_data_response <- mel_dat$meta.multiple.bin
test_data <- mel_dat
test_dat <- l_te_2
ann_model <- ann_model_1

  xgb_roc <- roc(test_data_response, 
                 predict(xgb_model, test_dat, type = "prob")[,1], 
                 levels = rev(levels(test_data_response)))
  rf_roc <- roc(test_data_response, 
                predict(rf_model, test_dat, type = "prob")[,1], 
                levels = rev(levels(test_data_response)))
  
  glmnet_roc <- roc(test_data_response, 
                    predict(glmnet_model, test_dat, type = "prob")[,1], 
                    levels = rev(levels(test_data_response)))
  
  test.hex <- as.h2o(test_data)
  pred <- h2o.predict(ann_model, test.hex)
  predicted_prob <- as.data.frame(pred)
  predicted_prob <- predicted_prob$High
  true_labels <- as.vector(test_data_response)
  ann_roc <- roc(response = true_labels, predictor = predicted_prob)
    

  
  

```


```{r,fig.height= 4, fig.width= 4}

plot.roc(ann_roc, las=1, lwd=1.7, col = "#386CB0", xlab=" 1 - specificity")

plot.roc(xgb_roc, add=TRUE, lwd=1.7, col = "#E41A1C")

plot.roc(rf_roc, add=TRUE, lwd=1.7, col = "#7FC97F")

plot.roc(glmnet_roc, add=TRUE, lwd=1.7, col =  "#999999")


legend("bottomright", legend = c(paste("Deep Learning", round(auc(ann_roc), 4)), 
                                 paste("XGBoost", round(auc(xgb_roc), 4)),
                                 paste("Random Forest", round(auc(rf_roc), 4)), 
                                 paste("GLMNet", round(auc(glmnet_roc), 4) )), 
                          lty = c(1, 1, 1, 1), 
                          col =c("#386CB0", "#E41A1C", "#7FC97F",  "#999999"), cex=0.6, lwd=1.5, inset = 0.05,
                                  title = "AUC")
                          par(mfrow = c(1, 1))



```

# Test Data Set AUC-ROC Plot


```{r,fig.height= 4, fig.width= 4}
test_data_response <- test$meta.multiple.bin
test_data <- test
test_dat <- l_te
ann_model <- ann_model_1

  xgb_roc <- roc(test_data_response, 
                 predict(xgb_model, test_dat, type = "prob")[,1], 
                 levels = rev(levels(test_data_response)))
  rf_roc <- roc(test_data_response, 
                predict(rf_model, test_dat, type = "prob")[,1], 
                levels = rev(levels(test_data_response)))
  
  glmnet_roc <- roc(test_data_response, 
                    predict(glmnet_model, test_dat, type = "prob")[,1], 
                    levels = rev(levels(test_data_response)))
  
  test.hex <- as.h2o(test_data)
  pred <- h2o.predict(ann_model, test.hex)
  predicted_prob <- as.data.frame(pred)
  predicted_prob <- predicted_prob$High
  true_labels <- as.vector(test_data_response)
  ann_roc <- roc(response = true_labels, predictor = predicted_prob)
  
  
```


```{r,fig.height= 4, fig.width= 4}
plot.roc(ann_roc, las=1, lwd=1.7, col = "#FB8072", xlab=" 1 - specificity")

plot.roc(xgb_roc, add=TRUE, lwd=1.7, col = "#80B1D3")

plot.roc(rf_roc, add=TRUE, lwd=1.7, col = "#B3DE69")

plot.roc(glmnet_roc, add=TRUE, lwd=1.7, col = "#FDB462")


legend("bottomright", legend = c(paste("Deep Learning", round(auc(ann_roc), 4)), 
                                 paste("XGBoost", round(auc(xgb_roc), 4)),
                                 paste("Random Forest", round(auc(rf_roc), 4)), 
                                 paste("GLMNet", round(auc(glmnet_roc), 4) )), 
                          lty = c(1, 1, 1, 1), 
                          col =c("#FB8072", "#80B1D3", "#B3DE69", "#FDB462"), cex=0.6, lwd=1.5, inset = 0.05,
                                  title = "AUC")
                          par(mfrow = c(1, 1))
```


```{r}
require(caret)
require(gridExtra)
require(h2o)

accuracy_table <- function(
    test_data,
    condition,
    xgb_model ,
    rf_model,
    # svmRadialWeights_model,
    glmnet_model,
    h2o_model
){
  models <- list(
    XGBoost = xgb_model,
    Random_Forest = rf_model,
    # svmRadialWeights_model = svmRadialWeights_model,
    GLMNet = glmnet_model,
    Deep_Learning = h2o_model
  )
  
  accuracy_results <- data.frame(
    Model = character(),
    Accuracy = numeric(),
    stringsAsFactors = FALSE
  )
  
  test_dat <- test_data[,colnames(test_data) != condition]
  
  for (model_name in names(models)) {
    model <- models[[model_name]]
    
    # Check if the model is an H2O model
    if (inherits(model, "H2OModel")) {
      # Predict using H2O model
      predictions <- as.factor(as.vector(h2o.predict(model, as.h2o(test_dat))$predict))
    } else {
      # Predict using non-H2O model
      predictions <- predict(model, test_dat)
    }
    
    # Calculate the confusion matrix and accuracy
    confusion <- confusionMatrix(predictions, test_data[[condition]])
    accuracy <- round(confusion$overall['Accuracy'], digits = 3)
    
    # Store the results in the data frame
    accuracy_results <- rbind(accuracy_results, data.frame(
      Model = model_name,
      Accuracy = accuracy
    ))
    
    rownames(accuracy_results) <- NULL
  }
  
  
  accuracy_results
  
  
  
}
```


# Test Data Set Accuracy Plot 

```{r,fig.height= 5, fig.width=4}

accuracy <- accuracy_table(
  test_data = test,
  condition = "meta.multiple.bin",
    xgb_model = xgb_model ,
    rf_model = rf_model,
    # svmRadialWeights_model,
    glmnet_model = glmnet_model,
    h2o_model = ann_model_1
  
)

palette <- c("#FB8072", "#80B1D3", "#B3DE69", "#FDB462")

names(palette) <- c("Deep_Learning","XGBoost", "Random_Forest", "GLMNet" )


library(ggpubr)
p <- ggbarplot(accuracy, x = "Model", y = "Accuracy", width = 0.5, fill = "Model", color = "black",
 label = TRUE, palette = palette, x.text.angle = 45)


p + theme(legend.position = "none")


```

# Validation Data Set Accuracy Plot 


```{r,fig.height= 5, fig.width=4}
accuracy <- accuracy_table(
  test_data = mel_dat,
  condition = "meta.multiple.bin",
    xgb_model = xgb_model ,
    rf_model = rf_model,
    # svmRadialWeights_model,
    glmnet_model = glmnet_model,
    h2o_model = ann_model_1
  
)

palette <- c("#386CB0", "#E41A1C", "#7FC97F",  "#999999")

names(palette) <- c("Deep_Learning","XGBoost", "Random_Forest", "GLMNet" )


library(ggpubr)
p <- ggbarplot(accuracy, x = "Model", y = "Accuracy", width = 0.5, fill = "Model", color = "black",
 label = TRUE, palette = palette, x.text.angle = 45)

p + theme(legend.position = "none")

#  "#7FC97F" "#BEAED4" "#FDC086" "#FFFF99" "#386CB0" "#F0027F" "#BF5B17" "#666666"
#  
#  [1] "#A6CEE3" "#1F78B4" "#B2DF8A" "#33A02C" "#FB9A99" "#E31A1C" "#FDBF6F" "#FF7F00" "#CAB2D6" "#6A3D9A"
# [11] "#FFFF99" "#B15928"
# 
#  
#  "#E41A1C" "#377EB8" "#4DAF4A" "#984EA3" "#FF7F00" "#FFFF33" "#A65628" "#F781BF" "#999999"
```






# important gene extraction

```{r}

#xgboost

xgbImp <- varImp(xgb_model, scale = T)
xgbImp_d <- data.frame(xgbImp$importance)
xgbImp_d <- xgbImp_d %>% filter(!Overall == 0)
xgbImp_d$gene <- row.names(xgbImp_d)
colnames(xgbImp_d)[colnames(xgbImp_d) == "Overall"] ="xgb_score"



# random forest

rfImp <- varImp(rf_model, scale = T)
rfImp_d <- data.frame(rfImp$importance)
rfImp_d <- rfImp_d  %>% filter(!Overall == 0)
rfImp_d$gene <- rownames(rfImp_d)
colnames(rfImp_d)[colnames(rfImp_d) == "Overall"] ="rf_score"
# rfImp_d$rf_score <- signif(rfImp_d$rf_score, digits  = 8) 

# glmnet

glmnetImp <- varImp(glmnet_model, scale = T)
glmnetImp_d <- data.frame(glmnetImp$importance)
glmnetImp_d <- glmnetImp_d %>% filter(!Overall == 0)
glmnetImp_d$gene <- rownames(glmnetImp_d)
colnames(glmnetImp_d)[colnames(glmnetImp_d) == "Overall"] ="glmnet_score"

# # svm
# 
# svmImp <- varImp(svmRadialWeights_model, scale = T)
# svmImp_d <- data.frame(svmImp$importance)
# svmImp_d <- svmImp_d %>% filter(!(High == 0 & Low == 0))

# identical(svmImp_d$High, svmImp_d$Low)
# [1] TRUE

# svmImp_d$gene <- rownames(svmImp_d)
# colnames(svmImp_d)[colnames(svmImp_d) == "High"] ="svm_score"
# svmImp_d <- svmImp_d %>% select(gene, svm_score)

# ann 

ann_varimp_1 <- h2o.varimp(ann_model_1)
ann_varimp_1_d <- as.data.frame(ann_varimp_1)
# identical(ann_varimp_1_d$relative_mportance, ann_varimp_1_d$scaled_mportance)
# [1] TRUE
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "variable"] ="gene"
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "scaled_mportance"] ="ann_score"
ann_varimp_1_d <- ann_varimp_1_d %>% select(gene, ann_score)

# diff expr.

test_diff_genes$significance<- "NO"
test_diff_genes$significance[test_diff_genes$padj < 0.05 & test_diff_genes$log2FoldChange > 1 ] <- "UP"
test_diff_genes$significance[test_diff_genes$padj < 0.05 & test_diff_genes$log2FoldChange < -1 ] <- "DOWN"

train_diff_genes$significance<- "NO"
train_diff_genes$significance[train_diff_genes$padj < 0.05 & train_diff_genes$log2FoldChange > 1 ] <- "UP"
train_diff_genes$significance[train_diff_genes$padj < 0.05 & train_diff_genes$log2FoldChange < -1 ] <- "DOWN"





test_diff_genes <- test_diff_genes %>% 
                   filter(significance != "NO") %>% 
                   mutate(wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, wlfc, significance, padj, log2FoldChange)


train_diff_genes <-train_diff_genes %>% 
                   filter(significance != "NO") %>% 
                   mutate(wlfc = -log2(padj)  * abs(log2FoldChange)) %>% 
                   select(gene, wlfc, significance, padj, log2FoldChange)

colnames(test_diff_genes) <- paste("test", colnames(test_diff_genes), sep = "_")
colnames(train_diff_genes) <- paste("train", colnames(train_diff_genes), sep = "_")

colnames(test_diff_genes)[colnames(test_diff_genes) == "test_gene"] ="gene"
colnames(train_diff_genes)[colnames(train_diff_genes) == "train_gene"] ="gene"


```











# Upsetr Plot for Downregulated Genes


```{r, fig.height=4, fig.width=8}

test_d <- test_diff_genes[test_diff_genes$test_significance == "DOWN", ]
train_d <- train_diff_genes[train_diff_genes$train_significance == "DOWN", ]


common_gene_list<- list(
  ann = ann_varimp_1_d$gene,
  # svm = svmImp_d$gene,
  xgb = xgbImp_d$gene,
  glmnet = glmnetImp_d$gene,
  train_diff_genes= test_d$gene,
  test_diff_genes = train_d $gene,
  rf = rfImp_d$gene
)

library(UpSetR)

# upset(fromList(common_gene_list), order.by = "freq")

common_elements <- Reduce(intersect, common_gene_list)

upset(fromList(common_gene_list), 
      nsets = 6, 
      order.by = "freq", 
      decreasing = T, 
      mb.ratio = c(0.55,0.45),
      number.angles = 0, 
      text.scale = 1.3,
      point.size = 2,
      line.size = 1,
      sets.bar.color = "red",
      main.bar.color = "darkgrey",
      mainbar.y.max = 2200,
      mainbar.y.label = "Common Gene Number",
      sets.x.label = "Important Gene Number"
      )



common_gene_list2 <- list(
  # ann = ann_varimp_1_d$gene,
  # svm = svmImp_d$gene,
  xgb = xgbImp_d$gene,
  glmnet = glmnetImp_d$gene,
  train_diff_genes= test_d$gene,
  test_diff_genes = train_d $gene,
  rf = rfImp_d$gene
)


upset(fromList(common_gene_list2), 
      nsets = 6, 
      order.by = "freq", 
      decreasing = T, 
      mb.ratio = c(0.55,0.45),
      number.angles = 0, 
      text.scale = 1.3,
      point.size = 2,
      line.size = 1,
      sets.bar.color = "red",
      main.bar.color = "darkgrey",
      mainbar.y.max = 2200,
      mainbar.y.label = "Common Gene Number",
      sets.x.label = "Important Gene Number"
      )

```

# Upsetr plot for upregulated genes


```{r, fig.height=4, fig.width=8}

test_d <- test_diff_genes[test_diff_genes$test_significance == "UP", ]
train_d <- train_diff_genes[train_diff_genes$train_significance == "UP", ]

common_gene_list<- list(
  ann = ann_varimp_1_d$gene,
  # svm = svmImp_d$gene,
  xgb = xgbImp_d$gene,
  glmnet = glmnetImp_d$gene,
  train_diff_genes= test_d$gene,
  test_diff_genes = train_d $gene,
  rf = rfImp_d$gene
)

library(UpSetR)

# upset(fromList(common_gene_list), order.by = "freq")

common_elements <- Reduce(intersect, common_gene_list)



upset(fromList(common_gene_list), 
      nsets = 6, 
      order.by = "freq", 
      decreasing = T, 
      mb.ratio = c(0.55,0.45),
      number.angles = 0, 
      text.scale = 1.3,
      point.size = 2,
      line.size = 1,
      sets.bar.color = "red",
      main.bar.color = "darkgrey",
      mainbar.y.max = 2200,
      mainbar.y.label = "Common Gene Number",
      sets.x.label = "Important Gene Number"
      )


common_gene_list2 <- list(
  # ann = ann_varimp_1_d$gene,
  # svm = svmImp_d$gene,
  xgb = xgbImp_d$gene,
  glmnet = glmnetImp_d$gene,
  train_diff_genes= test_d$gene,
  test_diff_genes = train_d $gene,
  rf = rfImp_d$gene
)


upset(fromList(common_gene_list2), 
      nsets = 6, 
      order.by = "freq", 
      decreasing = T, 
      mb.ratio = c(0.55,0.45),
      number.angles = 0, 
      text.scale = 1.3,
      point.size = 2,
      line.size = 1,
      sets.bar.color = "red",
      main.bar.color = "darkgrey",
      mainbar.y.max = 2200,
      mainbar.y.label = "Common Gene Number",
      sets.x.label = "Important Gene Number"
      )



# test_diff_genes <- select(test_diff_genes, -significance)
# train_diff_genes <- select(train_diff_genes, -significance)


```




```{r}
library(ggpubr)




expression_plotter <- function(expression_data, 
                               common_elements,
                               x_var,
                               x_tit
                               ){
plot_list <- list()

for (i in 1:length(common_elements)) {
  
gene <-  common_elements[i] 

p <- ggboxplot(expression_data, 
               x = x_var, 
               y = gene,
               xlab = x_tit,
               # color = "meta.multiple.bin", 
               palette = "jco",
               fill = x_var,
               add = "jitter", 
               shape = x_var)+ 
               stat_compare_means(method = "t.test", label.y = max(expression_data[[gene]])*1.2)+
               theme(legend.position = "none")

plot_list[[i]] = p
  
  
}

plot_list

}
```


# Box Plot for Common Elements  (Train Data Set was used )


```{r,fig.height= 3.2, fig.width=4, eval=FALSE}
plot_list <- expression_plotter(train, common_elements = common_elements, x_var = "meta.multiple.bin", x_tit = "Interferon Gamma Response")

plot_list

```

# Box Plot for Common Elements  (Test Data Set was used )

```{r,fig.height= 3.2, fig.width=4, eval=FALSE}
plot_list <- expression_plotter(test, common_elements = common_elements, x_var = "meta.multiple.bin", x_tit = "Interferon Gamma Response")

plot_list

```

# Box Plot for Up/Down Genes (Test Data Set was used )

```{r,fig.height= 3.2, fig.width=4}
gene_list = c("MS4A1", "CXCL10",  "PTPRN", "ADAM11", "EEF1A2", "WFDC1", "OCA2", "MAST1", "REEP6", "EFHD1"     
, "PAEP",  "MGAT5B", "CASKIN1", "LGI3", "CTXN1", "BEGAIN", "UBAP1L", "TPBGL")


plot_list <- expression_plotter(test, common_elements = gene_list, x_var = "meta.multiple.bin", x_tit = "Interferon Gamma Response")

plot_list
```

# Box Plot for Up/Down Genes (Train Data Set was used )

```{r,fig.height= 3.2, fig.width=4}
gene_list = c("MS4A1", "CXCL10",  "PTPRN", "ADAM11", "EEF1A2", "WFDC1", "OCA2", "MAST1", "REEP6", "EFHD1"     
, "PAEP",  "MGAT5B", "CASKIN1", "LGI3", "CTXN1", "BEGAIN", "UBAP1L", "TPBGL")


plot_list <- expression_plotter(train, common_elements = gene_list, x_var = "meta.multiple.bin", x_tit = "Interferon Gamma Response")

plot_list
```




```{r}
#xgboost

xgbImp <- varImp(xgb_model, scale = T)
xgbImp_d <- data.frame(xgbImp$importance)
# xgbImp_d <- xgbImp_d %>% filter(!Overall == 0)
xgbImp_d$gene <- row.names(xgbImp_d)
colnames(xgbImp_d)[colnames(xgbImp_d) == "Overall"] ="xgb_score"



# random forest

rfImp <- varImp(rf_model, scale = T)
rfImp_d <- data.frame(rfImp$importance)
# rfImp_d <- rfImp_d  %>% filter(!Overall == 0)
rfImp_d$gene <- rownames(rfImp_d)
colnames(rfImp_d)[colnames(rfImp_d) == "Overall"] ="rf_score"
# rfImp_d$rf_score <- signif(rfImp_d$rf_score, digits  = 8) 

# glmnet

glmnetImp <- varImp(glmnet_model, scale = T)
glmnetImp_d <- data.frame(glmnetImp$importance)
# glmnetImp_d <- glmnetImp_d %>% filter(!Overall == 0)
glmnetImp_d$gene <- rownames(glmnetImp_d)
colnames(glmnetImp_d)[colnames(glmnetImp_d) == "Overall"] ="glmnet_score"

# # svm
# 
# svmImp <- varImp(svmRadialWeights_model, scale = T)
# svmImp_d <- data.frame(svmImp$importance)
# svmImp_d <- svmImp_d %>% filter(!(High == 0 & Low == 0))

# identical(svmImp_d$High, svmImp_d$Low)
# [1] TRUE

# svmImp_d$gene <- rownames(svmImp_d)
# colnames(svmImp_d)[colnames(svmImp_d) == "High"] ="svm_score"
# svmImp_d <- svmImp_d %>% select(gene, svm_score)

# ann 

ann_varimp_1 <- h2o.varimp(ann_model_1)
ann_varimp_1_d <- as.data.frame(ann_varimp_1)
# identical(ann_varimp_1_d$relative_mportance, ann_varimp_1_d$scaled_mportance)
# [1] TRUE
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "variable"] ="gene"
colnames(ann_varimp_1_d)[colnames(ann_varimp_1_d) == "scaled_mportance"] ="ann_score"
ann_varimp_1_d <- ann_varimp_1_d %>% select(gene, ann_score)


```


# merging for distance

```{r}

v1 <- full_join(ann_varimp_1_d, rfImp_d, by= "gene") 
v2 <- full_join(xgbImp_d,glmnetImp_d, by= "gene")
v3 <- full_join(train_diff_genes,test_diff_genes, by= "gene")
v4 <- full_join(v1,v2, by= "gene")
v5 <- full_join(v3,v4, by= "gene")

merged_genes <- v5

rownames(merged_genes) <- merged_genes$gene

```


# common gene 

```{r, eval=FALSE}

common_data <- merged_genes %>% filter(gene %in% common_elements)

library(openxlsx)

write.xlsx(common_data, "common_table.xlsx")

```


```{r}

merged_genes <- select(merged_genes, -train_significance, -train_padj, -train_log2FoldChange, -test_significance, -test_padj, -test_log2FoldChange )

```




# data prep for corr plots

```{r}
mel_dat <- readRDS("validation_lcpm.rds")

l_dat <- readRDS("skcm_ready.RDS")

l_dat$meta.multiple.bin <- as.factor(l_dat$meta.multiple.bin)

l_dat <- l_dat %>% select(-meta.multiple.gene)

# l_dat <- l_dat %>% select(intersect(colnames(l_dat), colnames(mel_dat)))


library(caret)

set.seed(3456)

trainIndex <- createDataPartition(l_dat$meta.multiple.bin, p = .7, 
                                  list = FALSE, 
                                  times = 1)

train <- l_dat[ trainIndex,]

test <- l_dat[-trainIndex,]

summary(train$meta.multiple.bin)

summary(test$meta.multiple.bin)


l_tr <- train %>% select(-starts_with("meta."))
l_te <- test %>% select(-starts_with("meta."))

# mel_dat <- mel_dat %>% select(intersect(colnames(l_dat), colnames(mel_dat)))

l_te_2 <- mel_dat %>% select(-starts_with("meta."))
```


```{r, echo=FALSE, , message=FALSE}
train_data <- l_tr

library(purrr)
means <- data.frame(av_exp = map_dbl(train_data , mean))
means$gene <- rownames(means)

merged_data <- full_join(merged_genes, means, by="gene")

```







# Correlation Plots





## ann vs xgb

```{r,fig.height= 3, fig.width=4}

# "#A50026" "#D73027" "#F46D43" "#FDAE61" "#FEE090" "#FFFFBF" "#E0F3F8" "#ABD9E9" "#74ADD1" "#4575B4" "#313695"
# 
#  "#A6CEE3" "#1F78B4" "#B2DF8A" "#33A02C" "#FB9A99" "#E31A1C" "#FDBF6F" "#FF7F00" "#CAB2D6" "#6A3D9A"

library(ggpubr)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)

df <- merged_data

label = c("ESPTI1", "CXCL10", "APOL3", "LILRB2", "OASL")

p <- ggscatter(df, x = "ann_score", y = "xgb_score",
               xlab = "ANN Score",
               ylab = "XGBoost Score",
               fill = "av_exp",
               legend.position = "right",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21,
               ) +
  gradient_fill(c("#4575B4", "#FFFFBF", "#D73027"))+
  # labs(title = "ANN vs XGBoost")+
  theme(legend.position = "right")

# Add labels to outliers
# p + geom_text_repel(data = subset(df, ann_score > 0.68), aes(label = gene), vjust = -1, , color = "darkorange")

p + geom_label_repel(data = subset(df, ann_score > 0.95 | xgb_score > 90), aes(label = gene),  color = "black", fontface = "bold", size = 3, vjust = 0.8)



```

## ann vs rf

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

p <- ggscatter(df, x = "ann_score", y = "rf_score", 
               fill = "av_exp",
               xlab = "ANN Score",
               ylab = "Random Forest Score",
               size = 3,
               shape = 21,
               add.params = list(width = 0.2, height = 0.2)) +
  gradient_fill(c("#4575B4", "#FFFFBF", "#D73027"))+
  # labs(title = "ann_score vs rf_score")+
  theme(legend.position = "right")

# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene),  color = "black", fontface = "bold", size = 3)


# Add labels to outliers
p + geom_label_repel(data = subset(df, rf_score > 25 | ann_score > 0.95), aes(label = gene),color = "black", fontface = "bold", size = 3, hjust = 0.75)

```

## ann vs glmnet

```{r,fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "ann_score", y = "glmnet_score", 
               fill = "av_exp",
               xlab = "ANN Score",
               ylab = "Glmnet Score",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027"))+
  # labs(title = "ann_score vs glmnet_score")+
  theme(legend.position = "right")

# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "black",fontface = "bold",size = 3)


# Add labels to outliers
  p + geom_label_repel(data = subset(df, glmnet_score > 70 | ann_score > 0.90), aes(label = gene) ,  color = "black", fontface = "bold", size = 3, hjust = 0.8)

# p + geom_label_repel(data = subset(df, gene %in% label), aes(label = gene),  color = "black", fontface = "bold", size = 3)
```

## ann vs tr_padj

#### 15 common genes were labeled 

```{r, fig.height= 3, fig.width=4}
library(ggpubr)
library(ggplot2)

df <- merged_data
p <- ggscatter(df, y = "ann_score", x = "train_wlfc", 
               fill = "av_exp",
               ylab = "ANN Score",
               xlab = "P-Val Normalized l2fc (Train) ",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027"))+
  # labs(title = "ann_score vs glmnet_score")+
  theme(legend.position = "right")


# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "darkorange",fontface = "bold", size = 4.5)

p + geom_label_repel(data = subset(df, ann_score > 0.90 | train_wlfc > 1000) , aes(label = gene),  color = "black", fontface = "bold", size = 3, hjust = 0.8, vjust = 0.3)


```

## svm vs tr_padj

#### 15 common genes were labeled 

```{r, fig.height= 6, fig.width=8, eval=FALSE}
library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "svm_score", y = "train_wlfc", 
               color = "av_exp",
               size = 2,
               add.params = list(width = 0.2, height = 0.2)) +
  scale_color_gradient(low="blue", high="red")  +
  labs(title = "svm_score vs tr_wlp")+
  theme(legend.position = "right")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1.2, , color = "darkorange",fontface = "bold")


# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange")

```

## svm vs xgb

```{r, fig.height= 6, fig.width=8, eval=FALSE}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "svm_score", y = "xgb_score", 
               color = "av_exp",
               legend.position = "right",
               size = 3,
               add.params = list(width = 0.2, height = 0.2)) +
  scale_color_gradient(low="blue", high="red")  +
  labs(title = "svm_score vs xgb_score")+
  theme(legend.position = "right")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange",fontface = "bold")

# p + geom_text_repel(data = subset(df, xgb_score > 10), aes(label = gene), vjust = -1, , color = "darkorange")

```

## svm vs ann

```{r, fig.height= 6, fig.width=8, eval=FALSE}

p <- ggscatter(df, x = "svm_score", y = "ann_score", 
               color = "av_exp",
               size = 2,
               add.params = list(width = 0.2, height = 0.2)) +
  scale_color_gradient(low="blue", high="red")  +
  labs(title = "svm_score vs ann_score")+
  theme(legend.position = "right")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange",fontface = "bold")


# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange")

```

## svm vs rf

```{r, fig.height= 6, fig.width=8, eval=F}

library(ggpubr)
library(ggplot2)
library(ggrepel)

df <- merged_data

p <- ggscatter(df, x = "svm_score", y = "rf_score", 
               color = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2)) +
  scale_color_gradient(low="blue", high="red")  +
  labs(title = "svm_score vs rf_score")+
  theme(legend.position = "right")

# Add labels to outliers
# p + geom_text_repel(data = subset(df, rf_score > 70), aes(label = gene), vjust = -1, , color = "darkorange")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange",fontface = "bold")


```

## svm vs glmnet

```{r, fig.height= 3, fig.width=4, eval=FALSE}

library(ggpubr)
library(ggplot2)

df <- merged_data


p <- ggscatter(df, x = "svm_score", y = "glmnet_score", 
               color = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2)) +
  scale_color_gradient(low="blue", high="red")  +
  labs(title = "svm_score vs glmnet_score")+
  theme(legend.position = "right")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), vjust = -1, , color = "darkorange",fontface = "bold")


# Add labels to outliers
# p + geom_text_repel(data = subset(df, glmnet_score > 60), aes(label = gene), vjust = -1, , color = "darkorange")

```


## xgb vs glmnet

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "xgb_score", y = "glmnet_score", 
               fill = "av_exp",
               xlab = "XGBoost Score",
               ylab = "Glmnet Score",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027"))+
  # labs(title = "xgb_score vs glmnet_score")+
  theme(legend.position = "right")


# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene), color = "black",fontface = "bold", size = 3)


# Add labels to outliers
p + geom_text_repel(data = subset(df, glmnet_score > 65 |xgb_score > 25), aes(label = gene), color = "black",fontface = "bold", size = 3)

```

## xgb vs rf

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "xgb_score", y = "rf_score", 
               xlab = "XGBoost Score",
               ylab = "Random Forest",
               fill = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  # labs(title = "xgb_score vs rf_score")+
  theme(legend.position = "right")

# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "black",fontface = "bold", size = 3)


# Add labels to outliers
p + geom_label_repel(data = subset(df, rf_score > 25 |xgb_score > 25), aes(label = gene), color = "black",fontface = "bold", size = 3, hjust = 0.75)

```

## tr_padj vs xgb_score

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, y = "xgb_score", x = "train_wlfc", 
               fill = "av_exp",
               ylab = "XGBoost Score",
               xlab = "P-Val Normalized l2fc (Train) ",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  # labs(title = "xgb_score vs rf_score")+
  theme(legend.position = "right")

p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "black",fontface = "bold",size = 3, vjust = 1)


# Add labels to outliers
# p + geom_text_repel(data = subset(df, xgb_score > 75 |tr_wlp > 1800), aes(label = gene), vjust = -1, , color = "darkorange")

```

## tr_padj vs glmnet_score

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, y = "glmnet_score", x = "train_wlfc", 
               ylab = "Glmnet Score",
               xlab = "P-Val Normalized l2fc (Train) ",
               fill = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  # labs(title = "tr_wlp vs glmnet_score")+
  theme(legend.position = "right")

# p + geom_label_repel(data = subset(df, gene %in% label), aes(label = gene), color = "black",fontface = "bold",size = 3)


# Add labels to outliers
p + geom_label_repel(data = subset(df, glmnet_score > 50 |train_wlfc > 1100), aes(label = gene), color = "black",fontface = "bold",size = 3)

```

## tr_padj vs rf_score

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, y = "rf_score", x = "train_wlfc", 
               fill = "av_exp",
               ylab = "Random Forest Score",
               xlab = "P-Val Normalized l2fc (Train) ",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  # labs(title = "tr_wlp vs rf_score")+
  theme(legend.position = "right")

# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene)  , color = "black",fontface = "bold",size = 3, vjust = 1)


# Add labels to outliers
p + geom_text_repel(data = subset(df, rf_score > 25 |train_wlfc > 1100 ), aes(label = gene),, color = "black",fontface = "bold",size = 3, vjust = 1)

```


## glmnet_score vs rf_score

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "glmnet_score", y = "rf_score", 
               xlab = "Glmnet Score",
               ylab = "Random Forest Score ",
               fill = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  # labs(title = "tr_wlp vs rf_score")+
  theme(legend.position = "right")

# p + geom_text_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "black",fontface = "bold",size = 3, vjust = 1.5)


# Add labels to outliers
p + geom_text_repel(data = subset(df, glmnet_score > 80 |rf_score > 25), aes(label = gene), color = "black",fontface = "bold",size = 3, vjust = 1)

```

# te_wlp vs tr_wlp

```{r, fig.height= 3, fig.width=4}

library(ggpubr)
library(ggplot2)

df <- merged_data

p <- ggscatter(df, x = "train_wlfc", y = "test_wlfc", 
               ylab = "P-Val Normalized l2fc (Test)",
               xlab = "P-Val Normalized l2fc (Train) ",
               fill = "av_exp",
               size = 3,
               add.params = list(width = 0.2, height = 0.2),
               shape = 21
               
               ) +
 gradient_fill(c("#4575B4", "#FFFFBF", "#D73027")) +
  theme(legend.position = "right")


# p + geom_label_repel(data = subset(df, gene %in% label), aes(label = gene) , color = "black",fontface = "bold",size = 3)


# Add labels to outliers
p + geom_label_repel(data = subset(df, test_wlfc > 500 |train_wlfc > 1100), aes(label = gene)  , color = "black",fontface = "bold",size = 3)

```

```{r, eval=F}
v1 <- full_join(xgbImp_d,glmnetImp_d, by= "gene")
v2 <- full_join(train_diff_genes,test_diff_genes, by= "gene")
v3 <- full_join(v1,v2, by= "gene")
v4 <- full_join(rfImp_d,ann_varimp_1_d, by= "gene")
v5 <- full_join(v3,v4, by= "gene")

merged_genes <- v5

rownames(merged_genes) <- merged_genes$gene

merged_genes_scaled <- merged_genes %>% na.omit(merged_genes)

merged_genes_scaled  <- select(merged_genes_scaled, -gene)

```


```{r}
merged_genes_scaled <- merged_genes %>% na.omit(merged_genes)

merged_genes_scaled  <- select(merged_genes_scaled, -gene)
```



# Canberra Distance Matrix

```{r,fig.height= 4, fig.width=5.5}
library(pheatmap)
library(RColorBrewer)
library("FactoMineR")
library("devtools")
library("factoextra")
library(cluster)

distance_matrix_scaled <- as.matrix(get_dist(t(merged_genes_scaled), method = "canberra")) 


pheatmap(distance_matrix_scaled,
         main = "             Canberra Distance Matrix of Common Variables",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 

         color = rev(brewer.pal(n= 9, "RdYlBu"))
)
```


# Kendall Distance Matrix 

```{r,fig.height= 4, fig.width=5.5}
w_mat <- cor(merged_genes_scaled, method = 'kendall')

pheatmap(w_mat,
         main = "Kendall Matrix of Common Variables",
         # display_numbers = TRUE, 
         # fontsize_number = 10, 
         # annotation_col = alg_col,
         # annotation_row = alg_col,
         color = rev(brewer.pal(n= 9, "RdBu"))
)
```

```{r}

library(dplyr)

# file reading
tr_dif <- readRDS("TCGA_SKCM_train_de_genes_IFNG_final.rds") 
te_dif <- readRDS("TCGA_SKCM_test_de_genes_IFNG_final.rds")  

# removal of outlier genes

# tr_dif <- filter(tr_dif, ! is.na(padj))
# 
# te_dif <- filter(te_dif, ! is.na(padj))


```


# volcano plot for train data set

```{r, ,fig.height= 4, fig.width=5.5}
library(ggplot2)
library(ggrepel)

tr_dif$significance<- "NO"
tr_dif$significance[tr_dif$padj < 0.05 & tr_dif$log2FoldChange > 1 ] <- "UP"
tr_dif$significance[tr_dif$padj < 0.05 & tr_dif$log2FoldChange < -1 ] <- "DOWN"

label = common_elements

colors <- c("#bb190a", "#396ced", "#dce7e7")

names(colors) <- c("UP", "DOWN", "NO")

p <- ggplot(data=tr_dif, aes(x=log2FoldChange, y=-log10(padj), color = significance)) + 
  geom_point(size = 2,alpha = 0.2) + 
  theme_minimal()+
  scale_colour_manual(values = colors)+
  geom_hline(yintercept = -log10(0.05), color = "#b2bdbd")+
  geom_vline(xintercept = c(-1, 1), color = "#b2bdbd")+
  geom_text_repel(data =tr_dif %>% filter(tr_dif$gene %in% label),
            aes(label = gene),
            # vjust = -1,
            fontface = "bold",
            color = "black",
            size = 2.5)

p


```

# volcano plot for test data set

```{r, ,fig.height= 4, fig.width=5.5}
library(ggplot2)

te_dif$significance<- "NO"
te_dif$significance[te_dif$padj < 0.05 & te_dif$log2FoldChange > 1 ] <- "UP"
te_dif$significance[te_dif$padj < 0.05 & te_dif$log2FoldChange < -1 ] <- "DOWN"

# label = common_elements

colors <- c("#bb190a", "#396ced", "#dce7e7")

names(colors) <- c("UP", "DOWN", "NO")

p <- ggplot(data=te_dif, aes(x=log2FoldChange, y=-log10(padj), color = significance)) + 
  geom_point(size = 2,alpha = 0.2) + 
  theme_minimal()+
  scale_colour_manual(values = colors)+
  geom_hline(yintercept = -log10(0.05), color = "#b2bdbd")+
  geom_vline(xintercept = c(-1, 1), color = "#b2bdbd")+
  geom_text_repel(data =te_dif %>% filter(tr_dif$gene %in% label),
            aes(label = gene),
            # vjust = -1,
            fontface = "bold",
            color = "black",
            size = 2.5)


p


```