---
title: "This is Variable Importance Analysis for Different ML Algoritms"
author: Arda Askin
date: "`r Sys.Date()`"   
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true 
    code_folding: show
    theme: sandstone
--- 

TCGA SKCM Data was used to predict lymphocyte infiltration level. lymphocyte inflitration score was binarized as high and low. Before Diff. Exp. Analysis, the data was splitted test (n = 140) and train (n = 329)data sets. DSeq2 normalized data sets were transformed by Tukey's Transformation.(For more information about Tukey's Transformation: https://onlinestatbook.com/2/transformations/tukey.html) After Transformation, deeplearning, svm classifier, xgboost, randomforest and glmnet models were formed.    

# Model Performance 



```{r, echo=FALSE, , message=FALSE, , warning=FALSE}
library(dplyr)
library(caret)
library(MLeval)
library(pROC)
library(h2o)

h2o.init(nthreads = -1) 

```



```{r, echo=FALSE, , message=FALSE}
train_col_data <- readRDS("TCGA_SKCM_train_col_data.rds")

train_count_data <- readRDS("TCGA_SKCM_train_vsd.rds")

test_col_data <- readRDS("TCGA_SKCM_test_col_data.rds")

test_count_data <- readRDS("TCGA_SKCM_test_vsd.rds")

train_col_data$patients <- row.names(train_col_data)

train_count_data$patients <- row.names(train_count_data)

test_col_data$patients <- row.names(test_col_data)

test_count_data$patients <- row.names(test_count_data)

```

```{r, echo=FALSE}
train_data <- inner_join(train_count_data, train_col_data, by="patients")
train_data <- select(train_data, -patients)

test_data <- inner_join(test_count_data, test_col_data, by="patients")
test_data <- select(test_data, -patients)

train_data <- select(train_data, intersect(colnames(test_data), colnames(train_data)))
test_data <- select(test_data, intersect(colnames(test_data), colnames(train_data)))


```



```{r, echo=FALSE}
xgb_model <- readRDS("TCGA_SKCM_xgb_model_vsd.RDS")
rf_model<- readRDS("TCGA_SKCM_random_forest_model_vsd.rds")
svmRadialWeights_model <- readRDS("TCGA_SKCM_svmRadialWeights_model_vsd.rds")
glmnet_model <- readRDS("TCGA_SKCM_GLMNET_model_vsd.rds")

ann_model_1 <- h2o.loadModel("C:\\EKIZ_LAB\\ML_Performance\\Models\\TCGA_SKCM_model\\TCGA_SKCM_ann_model_vsd.rds\\dl_grid_model_1")


## diff expr.

train_diff_genes <- readRDS("TCGA_SKCM_train_de_genes.rds")
test_diff_genes <- readRDS("TCGA_SKCM_test_de_genes.rds")


```

## Performance Analysis 

### XGBOOST

```{r}

# eval metrics of train data

x_XGB <- evalm(xgb_model)


test_dat <- test_data %>% select(-meta.lymph.inf.sign)

xgb_roc <- roc(test_data$meta.lymph.inf.sign, 
            predict(xgb_model, test_dat, type = "prob")[,1], 
            levels = rev(levels(test_data$meta.lymph.inf.sign)))

plot(xgb_roc, print.auc=TRUE, 
     legacy.axes = TRUE)

xgb_predicted = predict(xgb_model, test_dat)

confusionMatrix(xgb_predicted, test_data$meta.lymph.inf.sign)

```

### Random Forest

```{r}
x_rf <- evalm(rf_model)

test_dat <- test_data %>% select(-meta.lymph.inf.sign)

library(pROC)

rf_roc <- roc(test_data$meta.lymph.inf.sign, 
            predict(rf_model, test_dat, type = "prob")[,1], 
            levels = rev(levels(test_data$meta.lymph.inf.sign)))

plot(rf_roc, print.auc=TRUE, 
     legacy.axes = TRUE)

rf_predicted = predict(rf_model, test_dat)

confusionMatrix(rf_predicted, test_data$meta.lymph.inf.sign)

xgb_predicted = predict(xgb_model, test_dat)

confusionMatrix(xgb_predicted, test_data$meta.lymph.inf.sign)

```

### GLMNET

```{r}

x_glmnet <- evalm(glmnet_model)

glmnet_roc <- roc(test_data$meta.lymph.inf.sign, 
            predict(glmnet_model, test_dat, type = "prob")[,1], 
            levels = rev(levels(test_data$meta.lymph.inf.sign)))

plot(glmnet_roc, print.auc=TRUE, 
     legacy.axes = TRUE)

glmnet_predicted = predict(glmnet_model, test_dat)

confusionMatrix(glmnet_predicted, test_data$meta.lymph.inf.sign)

```

### SVM Classifier

```{r}
x_svm <- evalm(svmRadialWeights_model)

svm_roc <- roc(test_data$meta.lymph.inf.sign, 
            predict(svmRadialWeights_model, test_dat, type = "prob")[,1], 
            levels = rev(levels(test_data$meta.lymph.inf.sign)))

plot(svm_roc, print.auc=TRUE, 
     legacy.axes = TRUE)

svm_predicted = predict(svmRadialWeights_model, test_dat)

confusionMatrix(svm_predicted, test_data$meta.lymph.inf.sign)


```

### Deep Learning 

```{r}
perf_1_cross <- h2o.performance(ann_model_1,  xval = TRUE)
 
plot(perf_1_cross, type = "roc")

perf_1_train <- h2o.performance(ann_model_1,  valid = TRUE)
 
plot(perf_1_train, type = "roc")

```


# Variable Importance

### XGBOOST

```{r}

xgbImp <- varImp(xgb_model, scale = T)
xgbImp_d <- data.frame(xgbImp$importance)
xgbImp_d <- xgbImp_d %>% filter(!Overall == 0)
xgbImp_d$genes <- row.names(xgbImp_d)

plot(xgbImp, top = 20, )

```


### Random Forest

```{r}

rfImp <- varImp(rf_model, scale = T)
rfImp_d <- data.frame(rfImp$importance)
rfImp_d <- rfImp_d  %>% filter(!Overall == 0)
rfImp_d$genes <- rownames(rfImp_d)

plot(rfImp, top = 20, )

```


### GLMNET

```{r}

glmnetImp <- varImp(glmnet_model, scale = T)
glmnetImp_d <- data.frame(glmnetImp$importance)
glmnetImp_d <- glmnetImp_d %>% filter(!Overall == 0)
glmnetImp_d$genes <- rownames(glmnetImp_d)

plot(glmnetImp, top = 20, )

```


### SVM Classifier

```{r}

svmImp <- varImp(svmRadialWeights_model, scale = T)
svmImp_d <- data.frame(svmImp$importance)
svmImp_d <- svmImp_d %>% filter(!(High == 0 & Low == 0))
svmImp_d$genes <- rownames(svmImp_d)


plot(svmImp, top = 20, )

```


### Deep Learning 

```{r}

source("h2o_imp_var.r")

ann_varimp_1 <- h2o.varimp(ann_model_1)
ann_varimp_1_d <- as.data.frame(ann_varimp_1)


h2o_imp_var(ann_model_1)



```



```{r}

## diff expr.

library(dplyr)

train_diff_high_up <- filter(train_diff_genes, log2FoldChange > 0 )
train_diff_high_down <- filter(train_diff_genes, !log2FoldChange > 0 )

test_diff_high_up <- filter(test_diff_genes, log2FoldChange > 0 )
test_diff_high_down <- filter(test_diff_genes, !log2FoldChange > 0 )

```

# Variable Importance Analysis

## Finding common important variables 

```{r}
ann_10 <- top_n(ann_varimp_1,round(nrow(ann_varimp_1)*0.1),relative_mportance)
svm_10 <- top_n(svmImp_d,round(nrow(svmImp_d)*0.1),High)

relative_ann <- ann_varimp_1_d %>% dplyr::filter(relative_mportance > median(relative_mportance))
relative_svm <- svmImp_d %>% dplyr::filter(High > median(High))

gene_list_up <- list(
  ann_10 = ann_10$variable,
  svm_10 = svm_10$genes, 
  train_diff_high_up = train_diff_high_up$gene,
  test_diff_high_up = test_diff_high_up$gene,
  # xgb = xgbImp_d$genes,
  rf = rfImp_d$genes
  # glmnet = glmnetImp_d$genes
)


library(UpSetR)

upset(fromList(gene_list_up), order.by = "freq")

common_elements_up <- Reduce(intersect, gene_list_up)
common_elements_up


gene_list_up_med <- list(
  relative_ann = relative_ann$variable,
  relative_svm = relative_svm$genes,
  train_diff_high_up = train_diff_high_up$gene,
  test_diff_high_up = test_diff_high_up$gene,
  # xgb = xgbImp_d$genes,
  rf = rfImp_d$genes
  # glmnet = glmnetImp_d$genes
)


upset(fromList(gene_list_up_med), order.by = "freq")

common_elements_up_med <- Reduce(intersect, gene_list_up_med)
common_elements_up_med

gene_list_down_med <- list(
  relative_ann = relative_ann$variable,
  relative_svm = relative_svm$genes,
  train_diff_high_down = train_diff_high_down$gene,
  test_diff_high_down = test_diff_high_down$gene,
  # xgb = xgbImp_d$genes,
  rf = rfImp_d$genes
  # glmnet = glmnetImp_d$genes
)

upset(fromList(gene_list_down_med), order.by = "freq")

common_elements_down_med <- Reduce(intersect, gene_list_down_med)
common_elements_down_med

gene_list_down <- list(
  ann_10 = ann_10$variable,
  svm_10 = svm_10$genes,
  train_diff_high_down = train_diff_high_down$gene,
  test_diff_high_down = test_diff_high_down$gene,
  # xgb = xgbImp_d$genes,
  rf = rfImp_d$genes
  # glmnet = glmnetImp_d$genes
)

upset(fromList(gene_list_down), order.by = "freq")

common_elements_down <- Reduce(intersect, gene_list_down)
common_elements_down

library(biomaRt)
 
mart <- useEnsembl("genes")
 
human <- useDataset("hsapiens_gene_ensembl", mart)
 
human_genes <- getBM(mart = human, attributes = c("hgnc_symbol", "external_gene_name", "gene_biotype"))
 
all_lncrna <- human_genes %>% dplyr::filter(gene_biotype =="lncRNA")

library("stringr") 

all_lncrna$gene_symb <- all_lncrna$external_gene_name

all_lncrna$external_gene_name <- str_replace_all( all_lncrna$external_gene_name,"-", ".")


linc_genes_high_up <- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_up )
linc_genes_high_up_med <- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_up_med )
linc_genes_high_down_med <- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_down_med)

linc_genes_high_down<- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_down)

linc_genes_train_down <- all_lncrna %>%dplyr::filter(external_gene_name %in% train_diff_high_down$gene)
linc_genes_test_down <- all_lncrna %>%dplyr::filter(external_gene_name %in% test_diff_high_down$gene)

linc_genes_train_up <- all_lncrna %>%dplyr::filter(external_gene_name %in% train_diff_high_up$gene)
linc_genes_test_up <- all_lncrna %>%dplyr::filter(external_gene_name %in% test_diff_high_up$gene)


```


```{r}

ann_10l <- top_n(ann_varimp_1,-round(nrow(ann_varimp_1)*0.1),relative_mportance)
svm_10l <- top_n(svmImp_d,-round(nrow(svmImp_d)*0.1),High)

gene_list_down_l <- list(
  ann_10l = ann_10l$variable,
  svm_10l = svm_10l$genes,
  train_diff_high_down = train_diff_high_down$gene,
  test_diff_high_down = test_diff_high_down$gene,
  rf = rfImp_d$genes
)

common_elements_down_l <- Reduce(intersect, gene_list_down_l)
common_elements_down_l

linc_genes_high_down_l<- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_down_l)

relative_annl <- ann_varimp_1_d %>% dplyr::filter(relative_mportance < median(relative_mportance))

relative_svml <- svmImp_d %>% dplyr::filter(High < median(High))

gene_list_down_med_l <- list(
  relative_annl = relative_annl$variable,
  relative_svml = relative_svml$genes,
  train_diff_high_down = train_diff_high_down$gene,
  test_diff_high_down = test_diff_high_down$gene,
  rf = rfImp_d$genes
)

upset(fromList(gene_list_down_med), order.by = "freq")

common_elements_down_medl <- Reduce(intersect, gene_list_down_med_l)
common_elements_down_medl

linc_genes_high_down_medl <- all_lncrna %>% dplyr::filter(external_gene_name %in% common_elements_down_medl)

```



## Examine the genes on TCGA SKCM Test Dataset 

```{r}

library(ggplot2)
library(ggpubr)



gene1 <-  "SPOCK2" 

ggboxplot(test_data, x = "meta.lymph.inf.sign", y = gene1,
                color = "meta.lymph.inf.sign", palette = "jco",
                add = "jitter", shape = "meta.lymph.inf.sign")+ 
            stat_compare_means(method = "t.test")



plot_list <- list()

for (i in 1:length(common_elements)) {
  
gene <-  common_elements[i] 

p <- ggboxplot(test_data, x = "meta.lymph.inf.sign", y = gene,
                color = "meta.lymph.inf.sign", palette = "jco",
                add = "jitter", shape = "meta.lymph.inf.sign")+ 
            stat_compare_means(method = "t.test")

plot_list[[i]] = p
  
  
}

plot_list


```


## Examine the genes on all-icb data set  


```{r}


external_data <- readRDS("all_icb.rds")

external_data <- external_data %>% filter(!is.na(external_data$meta.Response.y))

external_data <- external_data %>% 
  filter(meta.Response.y== "Partial Response" | meta.Response.y == "Complete Response"|meta.Response.y == "Progressive Disease") %>% 
  
  mutate(meta.immunotherapy.response = case_when(
    meta.Response.y == "Partial Response" |
    meta.Response.y == "Complete Response" ~ "R",
    meta.Response.y == "Progressive Disease" ~ "NR",
    
  ))

colnames(external_data) <- gsub("-", ".", colnames(external_data))

common_elements <- linc_genes_high_up_med$external_gene_name[-7]


# for all cancer type

plot_list <- list()

for (i in 1:length(common_elements)) {
  
gene <-  common_elements[i] 

p <- ggboxplot(external_data, x = "meta.immunotherapy.response", y = gene,
                color = "meta.immunotherapy.response", palette = "jco",
                add = "jitter", shape = "meta.immunotherapy.response")+ 
            stat_compare_means(method = "t.test")

p <- facet(p, facet.by = "meta.Timepoint")

plot_list[[i]] = p
  
  
}

plot_list

# for melanoma


external_data_melanoma <- external_data %>% filter(meta.Cancer_Tissue.y == "Skin")

plot_list_2 <- list()

for (i in 1:length(common_elements)) {
  
gene <-  common_elements[i] 

p <- ggboxplot(external_data_melanoma, x = "meta.immunotherapy.response", y = gene,
                color = "meta.immunotherapy.response", palette = "jco",
                add = "jitter", shape = "meta.immunotherapy.response")+ 
            stat_compare_means(method = "t.test")

p <- facet(p, facet.by = "meta.Timepoint")

plot_list_2[[i]] = p
  
  
}

plot_list_2




```


## examine importance score relationship

```{r}

xgb_vs_test_diff <- inner_join(xgbImp_d, test_diff_genes, by ="genes")

xgb_vs_train_diff <- inner_join(xgbImp_d, train_diff_genes, by = "genes")

xgb_vs_rf <- inner_join(xgbImp_d, rfImp_d, by = "genes")

xgb_vs_glmnet <- inner_join(xgbImp_d, glmnetImp_d, by= "genes")

xgb_vs_svm <- inner_join(xgbImp_d, svmImp_d, by= "genes")

rf_vs_test_diff <- inner_join(rfImp_d, test_diff_genes, by ="genes")

rf_vs_train_diff <- inner_join(rfImp_d, train_diff_genes, by = "genes")

rf_vs_glmnet <- inner_join(rfImp_d, glmnetImp_d, by= "genes")

rf_vs_svm <- inner_join(rfImp_d, svmImp_d, by= "genes")

glmnet_vs_test_diff <- inner_join(glmnetImp_d, test_diff_genes, by ="genes")

glmnet_vs_train_diff <- inner_join(glmnetImp_d, train_diff_genes, by = "genes")

glmnet_vs_svm <- inner_join(glmnetImp_d, svmImp_d, by= "genes")


```


```{r}
library(ggpubr)

df <- glmnet_vs_svm

x <- "Overall"

y <- "High"

xlab <- "glmnet_score"


ylab <- "svm_score"

title <- "glmnet_vs_svm"

ggscatter(df, x = x, y = y,
          add = "reg.line",                                 
          conf.int = FALSE, 
          # label = df$genes, 
          xlab = xlab,
          ylab = ylab,
          # repel = TRUE,
          title = title,
          add.params = list(color = "blue",
                            fill = "lightgray")
          )+
  stat_cor(method = "pearson", color = "red", label.x = 50, label.y = 50)  





```

